---
title: "Proposal"
author: "Maria Boyd"
date: "March 29, 2018"
output: html_document
---

#Introduction

This study will analyze five chemicals found in high concentrations on the Wolf River in Menominee County, Wisconsin. The five chemicals of interest are arsenic, chromium, phenanthrene, mercury, and manganese. Two of the five have been idenified as problematic by tribal members as recent talks about the Back40 mine ignited controversay. In an interivew Annie Wilber touches on existing impacts of manganese today and how detrimental it will become if the mine is approved. The potential for increased levels in thsee chemicals is astronomical. Interestingly testing has been discontinued since 2005 for the most prevalent chemicals identified except one, manganese.

2008 was the last year Menominee County was test for mercury. There are tests as far back as 1982 with records as high as 99.8 mg/l in 2003. Lead jump up to it's highest at 39 mg/l in downtown Menominee County. 

What toxic chemicals are in highest concentrations in Menominee County water?

What is the standard deviation for 3 chosen chemicals?
How many observations per chemical?

Of the three chemicals, which is most prevalent?
Are these three chemicals toxic to human health?

Last piece of information is going to pertain to remediation options or potential mitigation. Not sure I 

<!--chunk for getting data-->
```{r, RetreiveData}

#First input is a configuration of data identifying the highest level of pollution
#Retrieve function MC includes random characteristics
MC <-read.csv("MenomineeCounty.csv", stringsAsFactors = FALSE)
#head(MC)
tail(MC)

```

<!--chunk for year column in data frame-->
```{r yearcolumn}
#chunk converts dates to years
library(lubridate)
MC$Date <-ymd(MC$Date)
MC$Year <-year(MC$Date)
```

<!--chunk for Averages of each chemical-->
```{r AvChem}
AvChem <-data.frame(ChemicalNamec())

```


<!--chunk for filtering data set to only contain record of toxic chemicals we're interested in-->
```{r}

#Observations for all five chemicals
#Five chemicals arsenic, phenanthrene, mercury, manganese, chromium

MCAsMn <-MC[MC$ChemicalName%in%c("Manganese","Arsenic"),]
MCHgCoal <-MC[MC$ChemicalName%in%c("Phenanthrene","Mercury"),]
MCCr <-MC[MC$ChemicalName=="Chromium",]

#below function allows me to view chemicals in sets
View(MCAsMn)


```

<!--Outlines water quality standards for manganese toxicity to humans-->
[Tata & Howard Water Solution](https://tataandhoward.com/2017/01/importance-treating-manganese-drinking-water/)



```{r StandardDev}

#StandardDevCr <-sd("MCCr") THIS ONE DIDN'T WORK
Chemical_To_Analyze <- "Chromium"
StandardDevCr <-sd(MC[MC$ChemicalName==Chemical_To_Analyze,"Data"], na.rm = TRUE)#SD for 1984:2017

MeanCr <-mean(MC[MC$ChemicalName==Chemical_To_Analyze,"Data"], na.rm = TRUE)



CrSD <-aggregate(data~years,data=StandardDevCr)



```








